Fail Faster

### 2025-03-24 paper/paper.tex
remove commented cns
pdffig
cites

% Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
% \bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
% that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
% the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
% \bcp{This discussion should go elsewhere...}

@@

\paragraph{Staging eDSLs}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
\jwc{Cite racket macros for DSLs here:}
While embedded DSLs are powerful tools, they have a well-known drawback:
the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code.
This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
make eDSLs so pleasant to use are precisely what prevents them from being fast.
Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
not unique to generator DSLs, but pervasive across eDSLs.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
This recipe has been used to great effect to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, including
Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml \cite{metaocaml}) and Scala 3 (using
\texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml:
\scalaname{} is very similar.
For further discussion of the potential for staged generator DSLs in other languages, see Section~\ref{section:other-langs}.

@@

% carefully break up so that you push what you can into the compile time stage.
% Figuring out which parts of the generator are known statically ()
% Recall \jwc{did we talk about this} that staging a DSL involves changing the
% combinators to run at compile time by carefully annotating their types with \texttt{code}s.
% Deciding which types \texttt{t} can
% instead be \texttt{t code} --- in other words,
% figuring out which parts of the DSL can be determined statically (and can be
% part of the compile-time stage), and which parts are only known dynamically (and
% hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.

@@

The precise details of the SplitMix algorithm~\cite{splitmix} are not important for
present purposes; the key fact is that all of its operations are
defined in terms of arithmetic and bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer~\cite{ocamlruntimerep}. This means that all operations that return an \texttt{int64} must allocate this block of memory,
which has a significant impact on performance. A single call to
one of the \bq library functions---like generating an arbitrary integer---may call into the \texttt{Splittable\_random}
library times. Each call into \texttt{Splittable\_random}
may sample many times from the core SplitMix sampling routine \texttt{next\_int64}, i.e., using rejection sampling to find a value within a range.
Finally, each call to \texttt{next\_int64} allocates 9 times, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently
collect in OCaml\footnote{The OCaml GC is a generational collector~\cite{realworldocaml}, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}
, we will see in Section~\ref{section:eval} that they can still have
a large performance impact on generators that spend most of their time sampling data.
To avoid these allocations, we
reimplement SplitMix in C and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and
C code\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.}.

@@

All experiments were run on a 64-bit Linux machine with 264 GB RAM and a 128-core Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
\scalaname{} and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. For assessing and comparing PBT techniques, we used the Etna\cn platform.

@@

We select four strategies with disparate speedup profiles: insertion-based BSTs, single-pass BSTs,
boolean lists, and well-typed STLC terms. To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100 and
record the average number of \texttt{bind} calls. To determine the number of random samples, we repeat the same test but use Intel ProcessorTrace \cn{} to capture a 4ms snapshot of processor activity.
We then count the number of invocations of \texttt{next\_int64} and average this over the number of generator calls in the trace.

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking bug-finding speed}
To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
case studies using Etna\cn{}.
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test a specific aspect of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations).
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.
Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
expect the results in this section should extend to \scalaname.

@@

In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
in languages outside of the Haskell world from which it came.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
including non-strict or procedural ones --- might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.

@@

  \textbf{Racket} is the next target that we intend to test on.
  The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit. Indeed,
  there are at least two \cn{} PBT libraries in Racket.
  However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.

  \textbf{Haskell} is home to the original PBT implementation \cn{}, and
  so it is natural to ask why we have not built a \name{} in it yet. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
  complex enough programs however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system called Template Haskell \cn{}, which is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to build a \name{} version of Haskell
  QuickCheck, but the benefits would be much less pronounced.

  \textbf{F\#} has both multi-stage programming capabilities \cn{} and a well-used PBT library called FsCheck \cn{}.
  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building a \name{}
  in F\# is primarily an engineering effort.
  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations

  \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries \cn{}. However Rust does not
  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.

@@

\paragraph{Multi-Stage Programming}
% \jwc{A generator is a parser of
% randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
% lots of staged parser libraries!}
Staging's roots come from quasiquotation in LISP \cite{bawden99}, where the
unified representation of data and code allows for sophisticated
metaprogramming. Quasiquation-style macros were introduced to the ML family of
languages by Nielson and Nielson's Two-Level Functional Languages~\cn{} and
Taha's MetaML~\cn{}, which in turn quickly inspired MetaOCaml (and more recently
MacoCaml) in OCaml~\cn{}, Lightweight Modular Staging in Scala \cn{}, Template
Haskell in Haskell~\cn{}, and more.  In addition to implementations, researchers
have studied the type-theoretic foundations of multi-stage languages \cite{cmtt,
davies-pfenning, davies17}, as well as ways of combining multi-stage types with
other sophisticated features like dependent types \cite{kovacs24}. Meanwhile,
the Racket community continues the LISP tradition with a long and fruitful
line of work studying how macros and metaprogramming can be used to build extensible
DSLs \cite{flatt12, samth11, skthesis}.

@@

On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion \cn{} that yields optimal let placement,
which we did not use in \camlname{}.
Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions. Currently, the \camlname{}
recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this
overhead can be eliminated using a type-level heterogenous list to represent the state at compile time.

On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cn, WyHash~\cn,
and variants of xoroshiro~\cn. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
or even generating random numbers ahead of time. Last, we hope to investigate the degree to which actual statistical randomness
matters in PBT.

>>>>

% Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
% \bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
% that is ``split'' at \texttt{bind}s. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad to thread
% the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
% \bcp{This discussion should go elsewhere...}

@@

\paragraph{Staging eDSLs}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are powerful tools, they have a well-known drawback:
the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code.
This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
make eDSLs so pleasant to use are precisely what prevents them from being fast.
Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
not unique to generator DSLs, but pervasive across eDSLs.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
This recipe has been used to great effect to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, including
Scala 3~\cite{scalamacros}, Haskell~\cite{templatehaskell}, Racket~\cite{flatt12}, OCaml~\cite{metaocaml,macocaml}, and Java~\cite{mint}.
For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml~\cite{metaocaml}) and Scala 3 (using
\texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml:
\scalaname{} is very similar.
For further discussion of the potential for staged generator DSLs in other languages, see Section~\ref{section:other-langs}.

@@

% carefully break up so that you push what you can into the compile time stage.
% Figuring out which parts of the generator are known statically ()
% Recall \jwc{did we talk about this} that staging a DSL involves changing the
% combinators to run at compile time by carefully annotating their types with \texttt{code}s.
% Deciding which types \texttt{t} can
% instead be \texttt{t code} --- in other words,
% figuring out which parts of the DSL can be determined statically (and can be
% part of the compile-time stage), and which parts are only known dynamically (and
% hence must be \texttt{code}) --- is an art known as ``binding-time analysis''

@@

The precise details of the SplitMix algorithm~\cite{splitmix} are not important for
present purposes; the key fact is that all of its operations are
defined in terms of arithmetic and bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer~\cite{ocamlruntimerep}. This means that all operations that return an \texttt{int64} must allocate this block of memory,
which has a significant impact on performance. A single call to
one of the \bq library functions---like generating an arbitrary integer---may call into the \texttt{Splittable\_random}
library times. Each call into \texttt{Splittable\_random}
may sample many times from the core SplitMix sampling routine \texttt{next\_int64}, i.e., using rejection sampling to find a value within a range.
Finally, each call to \texttt{next\_int64} allocates 9 times, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently
collect in OCaml\footnote{The OCaml GC is a generational collector~\cite{realworldocaml}, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}
, we will see in Section~\ref{section:eval} that they can still have
a large performance impact on generators that spend most of their time sampling data.
To avoid these allocations, we
reimplement SplitMix in C and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and
C code\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types~\cite{unboxedtypes}, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.}.

@@

All experiments were run on a 64-bit Linux machine with 264 GB RAM and a 128-core Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
\scalaname{} and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16~\cite{corebench} in OCaml and \texttt{jmh} 0.4.7~\cite{jmh} for performance microbenchmarking. For assessing and comparing PBT techniques, we used the Etna platform~\cite{etna}.

@@

We select four strategies with disparate speedup profiles: insertion-based BSTs, single-pass BSTs,
boolean lists, and well-typed STLC terms. To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100 and
record the average number of \texttt{bind} calls. To determine the number of random samples, we repeat the same test but use Intel ProcessorTrace~\cite{intelpt} to capture a 4ms snapshot of processor activity.
We then count the number of invocations of \texttt{next\_int64} and average this over the number of generator calls in the trace.

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking bug-finding speed}
To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
case studies using Etna~\cite{etna}.
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test a specific aspect of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations).
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.
Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
expect the results in this section should extend to \scalaname.

@@

In the years since the original QuickCheck paper~\cite{quickcheck00}, PBT has had remarkable success
in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
including non-strict or procedural ones --- might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.

@@

  \textbf{Racket} is the next target that we intend to test on.
  The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it seems a natural fit.
  However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.

  \textbf{Haskell} is home to the original PBT implementation, and
  so it is natural to ask why we have not built a \name{} in it yet. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
  complex enough programs however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system called Template Haskell~\cite{templatehaskell}, which is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to build a \name{} version of Haskell
  QuickCheck, but the benefits would be much less pronounced.

  \textbf{F\#} has both multi-stage programming capabilities~\cite{fsmeta} and a well-used PBT library called FsCheck~\cite{fscheck}.
  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building an \name{}
  in F\# is primarily an engineering effort.
  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations

  \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.

@@

\paragraph{Multi-Stage Programming}
% \jwc{A generator is a parser of
% randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
% lots of staged parser libraries!}
Staging's roots come from quasiquotation in LISP \cite{bawden99}, where the
unified representation of data and code allows for sophisticated
metaprogramming. Quasiquation-style macros were introduced to the ML family of
languages by Nielson and Nielson's Two-Level Functional Languages~\cite{nielson92} and
MetaML~\cite{taha00}, which in turn quickly inspired MetaOCaml~\cite{metaocaml} (and more recently
MacoCaml~\cite{macocaml}) in OCaml, Lightweight Modular Staging in Scala~\cite{lms}, Template
Haskell in Haskell~\cite{templatehaskell}, and more.  In addition to implementations, researchers
have studied the type-theoretic foundations of multi-stage languages \cite{cmtt,
davies-pfenning, davies17}, as well as ways of combining multi-stage types with
other sophisticated features like dependent types \cite{kovacs24}. Meanwhile,
the Racket community continues the LISP tradition with a long and fruitful
line of work studying how macros and metaprogramming can be used to build extensible
DSLs \cite{flatt12, samth11, skthesis}.

@@

On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion~\cite{kiselyov2022letrecinsertioneffects} that yields optimal let placement,
which we did not use in \camlname{}.
Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions. Currently, the \camlname{}
recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this
overhead can be eliminated using a type-level heterogenous list to represent the state at compile time.

On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cite{lehmerrng}, WyHash~\cite{wyrand},
and variants of xoroshiro~\cite{xorshiro}. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
or even generating random numbers ahead of time. Last, we hope to investigate the degree to which actual statistical randomness
matters in PBT.
